\vspace{1.0cm} \LARGE \bf MERS-CoV spillover at the camel-human interface

### 2017-11-30 response_1/response_1.tex

\textbf{Yes, we agree that the structured coalescent approach is idealised and does not reflect a meaningful rate of zoonotic transfer of lineages, which is the reason we restricted any mention of rates to supplementary figures and do not attach numbers whenever rates are mentioned, but still report the number of introductions observed in the sequence data. We have altered Figure S2 in the reviewed manuscript to reflect that the rates shown are backwards in time. We have added the following sentences to the discussion to highlight the fact that the coalescent model employed is not ideal:}

\begin{quotation}
Although we recover migration rates from our model (Figure S2), these only pertain to sequences and in no way reflect the epidemiologically relevant \textit{per capita} rates of zoonotic spillover events.
\end{quotation}

\begin{quotation}
Having said that, there are highly customisable coalescent methods available that extend the methods used here to accommodate migration rates and population sizes varying through time, and integrate alternative sources of information that are able to fit stochastic noisy nonlinear models (Rasmussen \textit{et al.}., 2014) which would be more appropriate for MERS-CoV.
Some distinct aspects of MERS-CoV epidemiology could not be captured in our methodology, such as hospital outbreaks where $R_{0}$ is expected to be consistently closer to 1.0 compared to community transmission of MERS-CoV.
Outside of coalescent-based models there are population structure models that explicitly relate epidemiological parameters to the branching process observed in sequence data (K\"{u}hnert \textit{et al.}., 2016), but often rely on specifying numerous informative priors and can suffer from MCMC convergence issues.
\end{quotation}

@@

% \begin{quotation}
%
% \end{quotation}
2. p4, `Our analyses recover these results despite sequence data heavily skewed towards non-uniformly sampled human cases and are robust to choice of prior'. This is a quite nice result and raises the question if skewed sampling would bias estimates if using a substitution model approach (`discrete trait analysis', DTA). It would strengthen the paper to include a comparison of the structured coalescent estimates to another method for ancestral states; the most popular approach in beast has been substitution models (DTA). These may give divergent results because of skewed sampling. It would be rather easy for the authors to run a DTA and if biased, this would serve as a good cautionary example when sampling is highly skewed towards one deme.

\textbf{An excellent suggestion, thank you. We have run this analysis and include the results as a new supplementary figure S3. As expected, the skewed sampling results in a reconstruction of ancestral states that puts humans as the source of most MERS-CoV lineages in camels. We have added the appropriate description of methods as well as the following paragraph in results:}

\begin{quotation}
Our findings suggest that instances of human infection with MERS-CoV are more common than currently thought, with exceedingly short transmission chains mostly limited to primary cases that might be mild and ultimately not detected by surveillance or sequencing.
Structured coalescent analyses recover the camel-centered picture of MERS-CoV evolution despite sequence data heavily skewed towards non-uniformly sampled human cases and are robust to choice of prior.
Comparing these results with a currently standard discrete trait analysis (Lemey \textit{et al..}, 2009) approach for ancestral state reconstruction shows dramatic differences in host reconstruction at internal nodes (Figure S3).
Discrete trait analysis reconstruction identifies both camels and humans as important hosts for MERS-CoV persistence, but with humans as the ultimate source of camel infections.
A similar approach has been attempted previously (Zhang \textit{et al..}, 2016), but this interpretation of MERS-CoV evolution disagrees with lack of continuing human transmission chains outside of Arabian peninsula, low seroprevalence in humans and very high seroprevalence in camels across Saudi Arabia.
We suspect that this particular discrete trait analysis reconstruction is false due to biased data, \textit{i.e.} having nearly twice as many MERS-CoV sequences from humans (N=174) than from camels (N=100) and the inability of the model to account for and quantify vastly different rates of coalescence in the phylogenetic vicinity of both types of sequences.
\end{quotation}

@@

\textbf{A very good point. We share the suspicion that the results are largely driven by contrasts in effective population sizes between demes. In addition to the requested maximum likelihood phylogeny (supplementary Figure S14) we also ran a structured coalescent analysis where deme sizes are enforced to be the same (supplementary Figure S4). This model fails in a similar way to a DTA reconstruction shown in the new Figure S3. We now explain how the structured coalescent arrives at the tree shown in Figure 1 in the discussion:}

\begin{quotation}
When allowed different deme-specific population sizes, the structured coalescent model succeeds because a large proportion of human sequences fall into tightly connected clusters, which informs a low estimate for the population size of the human deme. 
This in turn informs the inferred state of long ancestral branches in the phylogeny, \textit{i.e.} because these long branches are not immediately coalescing, they are most likely in camels.
\end{quotation}

4. Easily addressed, but important. The paper already sounds a strong voice of concern in the final paragraph, but we think this could be even stronger. Antia et al. Nature 2003 first showed, using a simple branching process, that for most genetic landscapes, the probability of a pathogen evolving to state with R0>1 increases dramatically as a function of the wild-type R0. So R0~0.8 is much worse than R0~0.3. More sophisticated models have been done since, especially by Llyod-smith's group, but the basic result is sound. In the light of this theoretical work, your findings are not at all reassuring.

\textbf{We agree that this is an important point, but also feel that it is difficult to formulate warnings about pandemic potential without overstating the case. We also believe that adaptive landscapes play a considerable role in emerging pandemics. We added an additional sentence to the discussion and refer to the Antia et al. study:}

\begin{quotation}
Previous modeling studies (Antia \textit{et al.}., 2003; Part \textit{et al..}, 2013) suggest a positive relationship between initial $R_{0}$ in the human host and probability of evolutionary emergence of a novel strain which passes the supercritical threshold of $R_{0}>1.0$.
This leaves MERS-CoV in a precarious position; on one hand its current $R_{0}$ of $\sim$0.7 is certainly less than 1, but its proximity to the supercritical threshold raises alarm.
With very little known about the fitness landscape or adaptive potential of MERS-CoV, it is incredibly challenging to predict the likelihood of the emergence more transmissible variants.
In light of these difficulties, we encourage continued genomic surveillance of MERS-CoV in the camel reservoir and from sporadic human cases to rapidly identify a supercritical variant, if one does emerge.
\end{quotation}

@@

\textbf{The sentence was meant to summarise our findings. We have changed the sentence to clarify that the statement is based on our findings:}

\begin{quotation}
However, we find that MERS-CoV, once introduced into humans, follows transient transmission chains that soon abate.
\end{quotation}

2. In discussion: Could the author speculate how more sequences from humans and animals, particularly recent ones, or sequences from other geographic areas (east africa, north africa, pakistan) would potentially impace their results?

\textbf{We agree that understanding the distribution of MERS zoonotic risk to humans is of paramout importance, but feel that such speculation is outside the remit of the current study.
During the time the study was performed (first half of 2017) human MERS-CoV sequences comprised a large majority of available MERS-CoV sequences.
A study published during the first round of review (November 2017) that contributed over 100 new MERS-CoV genomes from camels in United Arab Emirates (Yusof \textit{et al..}, 2017, doi:10.1038/emi.2017.89) did not uncover genetic diversity that had not been seen in the Arabian peninsula previously.
We are also aware of another study into camels across much of Africa.
These are all consistent with general statements we have made in the manuscript in the methods section so far, which we have expanded further in response to this request:}

@@

Were more sequences of MERS-CoV available from other parts of Africa we speculate they would fall outside of the diversity that has been sampled in Saudi Arabia and cluster with early MERS-CoV sequences from Jordan and sequences from Egyptian camels.
However, currently there are no indications of what MERS-CoV diversity looks like in camels east of Saudi Arabia.
\end{quotation}

\textbf{We also intend to eventually update the manuscript with a structured coalescent analysis that includes all the sequence data that were made available recently.}

\section*{Reviewer \#3}

1. Page 2, introduction. I am not sure I would refer to the Cauchemez et al. PNAS paper as traditional epidemiology. It's sophisticated statistical modelling of spatio-temporal data and the partial observation process. I think you are using different tools in epidemiology, obtaining complimentary insights, neither particularly traditional. You could just delete `traditional' or use `case-based'.

@@

\begin{quotation}
Another hallmark of recombination is clustering of alleles along the genome, due to how template switching, the primary mechanism of recombination in RNA viruses, occurs.
3Seq relies on the clustering of nucleotide similarities along the genome between sequence triplets -- two potential parent-donors and one candidate offspring-recipient sequences.
\end{quotation}

7. Page 9. I think the section on recombination is verbose, and potentially misleading. You didn't perform an ARG, you performed a perfectly reasonable sensitivity analysis to different partitions of the genome. I would be more upfront about this, and try to explain it more simply and clearly.

\textbf{We have tried our best to clean up the section on recombination.}

8. Discussion. I think the results of this analysis on R0 and cluster size distribution could be compared with Cauchemez et al. PNAS 2016, who estimated many of the similar quantities using spatiotemporal data to define clusters. An important result of that paper was that the cluster size distributions were very skew - I think it would be good to look at skewness here.

\textbf{An excellent suggestion, however the authors of the Cauchemez \textit{et al.}. study did not share any data beyond the raw input file they used. Since the actual numbers in Fig 1C in the manuscript are not reported and a publicly available implementation of the MCMC algorithm do not exist we have included an additional figure that visualises the distribution of cases recovered by simulation. These exhibit a heavy left skew, much like the distribution in Cauchemez \textit{et al.}.}

### 2017-11-30 response_1/response_1.tex

% \begin{quotation}
%
% \end{quotation}
2. p4, `Our analyses recover these results despite sequence data heavily skewed towards non-uniformly sampled human cases and are robust to choice of prior'. This is a quite nice result and raises the question if skewed sampling would bias estimates if using a substitution model approach (`discrete trait analysis', DTA). It would strengthen the paper to include a comparison of the structured coalescent estimates to another method for ancestral states; the most popular approach in beast has been substitution models (DTA). These may give divergent results because of skewed sampling. It would be rather easy for the authors to run a DTA and if biased, this would serve as a good cautionary example when sampling is highly skewed towards one deme.

\textbf{An excellent suggestion, thank you. We have run this analysis and include the results as a new supplementary figure S3. As expected, the skewed sampling results in a reconstruction of ancestral states that puts humans as the source of most MERS-CoV lineages in camels. We have added the appropriate description of methods as well as the following paragraph in results:}

\begin{quotation}
Our findings suggest that instances of human infection with MERS-CoV are more common than currently thought, with exceedingly short transmission chains mostly limited to primary cases that might be mild and ultimately not detected by surveillance or sequencing.
Structured coalescent analyses recover the camel-centered picture of MERS-CoV evolution despite sequence data heavily skewed towards non-uniformly sampled human cases and are robust to choice of prior.
Comparing these results with a currently standard discrete trait analysis (Lemey \textit{et al.}, 2009) approach for ancestral state reconstruction shows dramatic differences in host reconstruction at internal nodes (Figure S3).
Discrete trait analysis reconstruction identifies both camels and humans as important hosts for MERS-CoV persistence, but with humans as the ultimate source of camel infections.
A similar approach has been attempted previously (Zhang \textit{et al.}, 2016), but this interpretation of MERS-CoV evolution disagrees with lack of continuing human transmission chains outside of Arabian peninsula, low seroprevalence in humans and very high seroprevalence in camels across Saudi Arabia.
We suspect that this particular discrete trait analysis reconstruction is false due to biased data, \textit{i.e.} having nearly twice as many MERS-CoV sequences from humans (N=174) than from camels (N=100) and the inability of the model to account for and quantify vastly different rates of coalescence in the phylogenetic vicinity of both types of sequences.
\end{quotation}

@@

\begin{quotation}
When allowed different deme-specific population sizes, the structured coalescent model succeeds because a large proportion of human sequences fall into tightly connected clusters, which informs a low estimate for the population size of the human deme. 
This in turn informs the inferred state of long ancestral branches in the phylogeny, \textit{i.e.} because these long branches are not immediately coalescing, they are most likely in camels.
\end{quotation}

4. Easily addressed, but important. The paper already sounds a strong voice of concern in the final paragraph, but we think this could be even stronger. Antia et al. Nature 2003 first showed, using a simple branching process, that for most genetic landscapes, the probability of a pathogen evolving to state with R0>1 increases dramatically as a function of the wild-type R0. So R0~0.8 is much worse than R0~0.3. More sophisticated models have been done since, especially by Llyod-smith's group, but the basic result is sound. In the light of this theoretical work, your findings are not at all reassuring.

\textbf{We agree that this is an important point, but also feel that it is difficult to formulate warnings about pandemic potential without overstating the case. We also believe that adaptive landscapes play a considerable role in emerging pandemics. We added an additional sentence to the discussion and refer to the Antia et al. study:}

\begin{quotation}
Previous modeling studies (Antia \textit{et al.}., 2003; Part \textit{et al.}, 2013) suggest a positive relationship between initial $R_{0}$ in the human host and probability of evolutionary emergence of a novel strain which passes the supercritical threshold of $R_{0}>1.0$.
This leaves MERS-CoV in a precarious position; on one hand its current $R_{0}$ of $\sim$0.7 is certainly less than 1, but its proximity to the supercritical threshold raises alarm.
With very little known about the fitness landscape or adaptive potential of MERS-CoV, it is incredibly challenging to predict the likelihood of the emergence more transmissible variants.
In light of these difficulties, we encourage continued genomic surveillance of MERS-CoV in the camel reservoir and from sporadic human cases to rapidly identify a supercritical variant, if one does emerge.
\end{quotation}

@@

\textbf{The sentence was meant to summarise our findings. We have changed the sentence to clarify that the statement is based on our findings:}

\begin{quotation}
However, we find that MERS-CoV, once introduced into humans, follows transient transmission chains that soon abate.
\end{quotation}

2. In discussion: Could the author speculate how more sequences from humans and animals, particularly recent ones, or sequences from other geographic areas (east africa, north africa, pakistan) would potentially impace their results?

\textbf{We agree that understanding the distribution of MERS zoonotic risk to humans is of paramout importance, but feel that such speculation is outside the remit of the current study.
During the time the study was performed (first half of 2017) human MERS-CoV sequences comprised a large majority of available MERS-CoV sequences.
A study published during the first round of review (November 2017) that contributed over 100 new MERS-CoV genomes from camels in United Arab Emirates (Yusof \textit{et al.}, 2017, doi:10.1038/emi.2017.89) did not uncover genetic diversity that had not been seen in the Arabian peninsula previously.
We are also aware of another study into camels across much of Africa.
These are all consistent with general statements we have made in the manuscript in the methods section so far, which we have expanded further in response to this request:}

### 2017-11-30 response_1/response_1.tex

\begin{quotation}
When allowed different deme-specific population sizes, the structured coalescent model succeeds because a large proportion of human sequences fall into tightly connected clusters, which informs a low estimate for the population size of the human deme. 
This in turn informs the inferred state of long ancestral branches in the phylogeny, \textit{i.e.} because these long branches are not immediately coalescing, they are most likely in camels.
\end{quotation}

4. Easily addressed, but important. The paper already sounds a strong voice of concern in the final paragraph, but we think this could be even stronger. Antia et al. Nature 2003 first showed, using a simple branching process, that for most genetic landscapes, the probability of a pathogen evolving to state with R0>1 increases dramatically as a function of the wild-type R0. So R0~0.8 is much worse than R0~0.3. More sophisticated models have been done since, especially by Llyod-smith's group, but the basic result is sound. In the light of this theoretical work, your findings are not at all reassuring.

\textbf{We agree that this is an important point, but also feel that it is difficult to formulate warnings about pandemic potential without overstating the case. We also believe that adaptive landscapes play a considerable role in emerging pandemics. We added an additional sentence to the discussion and refer to the Antia et al. study:}

\begin{quotation}
Previous modeling studies (Antia \textit{et al}., 2003; Part \textit{et al.}, 2013) suggest a positive relationship between initial $R_{0}$ in the human host and probability of evolutionary emergence of a novel strain which passes the supercritical threshold of $R_{0}>1.0$.
This leaves MERS-CoV in a precarious position; on one hand its current $R_{0}$ of $\sim$0.7 is certainly less than 1, but its proximity to the supercritical threshold raises alarm.
With very little known about the fitness landscape or adaptive potential of MERS-CoV, it is incredibly challenging to predict the likelihood of the emergence more transmissible variants.
In light of these difficulties, we encourage continued genomic surveillance of MERS-CoV in the camel reservoir and from sporadic human cases to rapidly identify a supercritical variant, if one does emerge.
\end{quotation}

@@

\textbf{We agree that the simulation study could have been explained better. We elaborated the description of simulations in results to say:}

\begin{quotation}
Sequencing simulations are performed via a multivariate hypergeometric distribution, where the probability of sequencing from a particular cluster depends on available sequencing capacity (number of trials), numbers of cases in the cluster (number of successes) and number of cases outside the cluster (number of failures).
In addition, each hypergeometric distribution used to simulate sequencing is concentrated via a bias parameter, that enriches for large sequence clusters at the expense of smaller ones.
\end{quotation}

3. p 8-11, What are the implications of recombinant MERS CoV in camels for estimation of evolutionary rates and TMRCAs?

\textbf{An excellent point. Concerns over artefacts of recombination are what prevented us from reporting on evolutionary rates and tMRCAs of interest, such as our postulated introduction of the virus into the Arabian peninsula. We added a discussion about the potential effects of recombination in the results section and report evolutionary rates recovered by two methods:}

### 2017-12-01 mers-structure.tex
not sure what changed, sorry...

Despite lack of convergence, neither of the two demographic reconstructions show evidence of fluctuations in the relative genetic diversity ($N_e \tau$) of  MERS-CoV over time (Figure \ref{skygrid}).
We recover a similar demographic trajectory when estimating $N_{e}\tau$ over time with a skygrid tree prior across the genome split into ten fragments with independent phylogenetic trees to account for confounding effects of recombination (Figures \ref{skygrid}, \ref{skygrid_comparison}).
However, we do note that coalescence rate estimates are high relative to the sampling time period, with a mean estimate of $N_e\tau$ at 3.49 years (95\% HPD: 2.71--4.38), and consequently MERS-CoV phylogeny resembles a ladder often seen in human influenza A virus phylogenies \citep{bedford_strength_2011}.

>>>>

Despite lack of convergence, neither of the two demographic reconstructions show evidence of fluctuations in the relative genetic diversity ($N_e \tau$) of  MERS-CoV over time (Figure \ref{skygrid}).
We recover a similar demographic trajectory when estimating $N_{e}\tau$ over time with a skygrid tree prior across the genome split into ten fragments with independent phylogenetic trees to account for confounding effects of recombination (Figures \ref{skygrid}, \ref{skygrid_comparison}).
However, we do note that coalescence rate estimates are high relative to the sampling time period, with a mean estimate of $N_e\tau$ at 3.49 years (95\% HPD: 2.71--4.38), and consequently MERS-CoV phylogeny resembles a ladder, as often seen in human influenza A virus phylogenies \citep{bedford_strength_2011}.
